An Intrinsically-Typed Solution for the\\ List-Machine Benchmark

### 2021-11-26 papers/COLA/paper.tex
finishing review for COLA

\begin{abstract}
Formal models are important tools in the programming language research
community. However, such models are full of intricacies and, due to that,
they are subject to subtle errors. Such failures motivated the usage of
tools to ensure the correctness of these formalisms. One way to eliminate
such errors is to encode models in a dependently-typed language in order
to ensure its ``correctness-by-construction''. In this paper, we use this
idea to build a verified interpreter for the list-machine benchmark in the
Agda programming language, comparing the results with formalizations developed
by Appel et. al. We formalize the 14 tasks of the benchmark
using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
compared to a Coq solution, even without the use of proof automation. We also
describe a solution to the second version of the benchmark and compare it with
Appel'et. al. Coq-based solution.
\end{abstract}

@@

In this work, we extend our previous SBLP 2020 paper~\cite{Feitosa2020} by including support to
indirect jumps, as specified in the version 2.0 of the list
machine benchmark~\cite{AppelDL12}. We provide detailed explanations of the needed
changes on the machine syntax, type system and interpreter to accomodate this extension, which
allows the handling of more complex control flow structures, like the call/return model of function calls.

@@

\begin{itemize}
  \item We show how all the details of the list-machine type system
        can be encoded in a dependently-typed syntax setting, which avoids, by construction,
        the presence of stuck states in its definitional interpreter.
  \item We provide a provably correct implementation for testing the subtyping
        relation and to calculate the least common supertype of two input
        types for the machine registers.
  \item We provide a correct-by-construction implementation of a type-checker for
        list-machine programs. Our type-checker produces, as a result, an intrinsically-typed
        representation of the machine code.
  \item We compare the results of an intrinsic approach encoded in Agda with extrinsic
        formalizations encoded in Twelf and Coq. We show that such intrinsic encoding avoids
        unnecessary repetitions and provides some properties for free.
  \item We provide a detailed discussion on the necessary modifications to the type system,
        type checker and interpreter to support indirect jumps.
\end{itemize}

@@

Dependently typed pattern matching is built by using the so-called
\ensuremath{\mathkw{with}} construct, that allows for matching intermediate
values~\cite{McBride2004}. If the matched value has a dependent type,
then its result can affect the form of other values. For example,
consider the following code that defines a type for natural number
parity. If the natural number is even, it can be represented as the
sum of two equal natural numbers; if it is odd, it is equal to one
plus the sum of two equal values. Pattern matching on a value of
\ensuremath{\D{Parity}\;\V{n}} allows to discover if $n = j + j$ or $n = S (k + k)$,
for some $j$ and $k$ in each branch of \ensuremath{\mathkw{with}}.  Note that the
value of $n$ is specialized accordingly, using information ``learned''
by the type-checker.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
\column{43}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathkw{data}\;\D{Parity}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\Con{Even}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\;\mathbin{:}\;\D{\mathbb{N}}\mskip1.5mu\}\;\to \;\D{Parity}\;(\V{n}\;\F{+}\;\V{n}){}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\Con{Odd}\;{}\<[9]%
\>[9]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\;\mathbin{:}\;\D{\mathbb{N}}\mskip1.5mu\}\;\to \;\D{Parity}\;(\V{S}\;(\V{n}\;\F{+}\;\V{n})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\F{parity}\;\mathbin{:}\;(\V{n}\;\mathbin{:}\;\D{\mathbb{N}})\;\to \;\D{Parity}\;\V{n}{}\<[E]%
\\
\>[B]{}\F{parity}\;\mathrel{=}\orange{\texttt{-- definition omitted}}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\F{natToBin}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{List}\;\D{Bool}{}\<[E]%
\\
\>[B]{}\F{natToBin}\;\Con{zero}\;\mathrel{=}\;\Con{\lbrack\:\rbrack}{}\<[E]%
\\
\>[B]{}\F{natToBin}\;\V{k}\;\mathkw{with}\;(\F{parity}\;\V{k}){}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\F{natToBin}\;(\V{j}\;\F{+}\;\V{j})\;{}\<[25]%
\>[25]{}\mid \;\Con{Even}\;\mathrel{=}\;\Con{false}\;\Con{::}\;\F{natToBin}\;\V{j}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\F{natToBin}\;(\Con{suc}\;(\V{j}\;\F{+}\;\V{j}))\;\mid \;\Con{Odd}\;{}\<[35]%
\>[35]{}\mathrel{=}\;\Con{true}\;{}\<[43]%
\>[43]{}\Con{::}\;\F{natToBin}\;\V{j}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
A important application of Agda's dependently-typed pattern matching is
to rewrite a equality. As an example, consider the following code which
postulates the commutativity of natural number addition and a predicate
over natural numbers, \ensuremath{\D{P}}.

@@

An important application of dependent types is to encode programming languages
syntax. The role of dependent types in this domain is to encode programs that
only allow well-typed and well-scoped terms~\cite{Benton2012}. Intuitively, we encode
the static semantics of the object language in the host language AST's
constructor, leaving the responsibility of checking type safety to the
host's language type-checker. Before considering our solution to encode list-machine
programs as instrisincally-typed syntax, we review this approach in a reduced example.

@@

Each program variable is assigned to a list type (Figure~\ref{fig:type-syntax}), which is used to guarantee the safety of executing
fetch-field instructions that demands non-empty list arguments. In order to express such refinements, types are subject to a
subtyping relation. The meta-variable $\tau$ denotes an arbitrary type.
\begin{figure}[H]
\[
\begin{array}{rcll}
  \tau & ::=  & \text{nil} & \text{(type for empty lists)}\\
       & \mid & \text{list }\tau & \text{(lists whose elements have type $\tau$)}\\
       & \mid & \text{listcons }\tau & \text{(non-empty lists of $\tau$)}\\
\end{array}
\]
\centering
\caption{Type syntax}
\label{fig:type-syntax}
\end{figure}
The notation $\tau \subset \tau'$
denotes the subtyping judgment, which is defined as follows.
\begin{figure}[H]
\[
\begin{array}{c}
  \inference{}
            {\tau\subset \tau}
            [subtype-refl]
  \\ \\
  \inference{}
            {nil \subset list\:\tau}
            [subtype-nil]
  \\ \\
  \inference{\tau \subset \tau'}
            {list\:\tau\subset list\:\tau'}
            [subtype-list]\\ \\
  \multicolumn{3}{c}{
  \inference{\tau \subset \tau'}
            {listcons\:\tau\subset list\:\tau'}
            [subtype-listmixed]} \\ \\
  \multicolumn{3}{c}{
            \inference{\tau \subset \tau'}
            {listcons\:\tau\subset listcons\:\tau'}
            [subtype-listcons]} \\ \\
\end{array}
\]
\centering
\caption{Subtyping relation}
\label{fig:subtyping}
\end{figure}
Basically, the subtyping relation specifies that $nil$ (empty list type) is
subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
The other rules specify that type constructors $list$ and $listcons$ respect
the subtyping relation. The least common supertype $\tau = \tau_1 \sqcap \tau_2$ of
$\tau_1$ and $\tau_2$ is defined as the smallest type such that $\tau_1$ and $\tau_2$
are subtypes of $\tau$.

@@

\section{Intrinsically-typed syntax}\label{sec:typing}

@@

\>[26]{}\rightarrow\;\V{\tau_1}\;\D{\in}\;\V{\Gamma}\;\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcup}\;\V{\tau_1}\;\D{\sim}\;\Con{list}\;\V{\tau}\;{}\<[E]%

@@

\>[26]{}\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcup}\;\V{\tau_1}\;\D{\sim}\;\Con{list}\;\V{\tau}\;{}\<[E]%

@@

\paragraph{List construction}{The \ensuremath{\Con{instr\textrm{-}cons\textrm{-}new}} and \ensuremath{\Con{instr\textrm{-}cons\textrm{-}upd}} constructors are
used to create a new list. The first creates a new variable, and the second updates an
existing variable. The list is created from two variables, \ensuremath{\V{\tau_0}\;\D{\in}\;\V{\Gamma}} and \ensuremath{\V{\tau_1}\;\D{\in}\;\V{\Gamma}},
which are represented as \emph{de Bruijn} indices. The type of the new list is defined by
the least common supertype\footnote{A complete explanation about the least common supertype
can be found in the original list-machine paper~\cite{Appel07}.}, which is defined by the
type \ensuremath{\V{\tau_1}\;\D{\sqcup}\;\V{\tau_2}\;\D{\sim}\;\V{\tau}}, which encodes that the least common supertype of \ensuremath{\V{\tau_1}} and \ensuremath{\V{\tau_2}} is \ensuremath{\V{\tau}}.
The resulting typing context adds information about the newly created list.
Subsection~\ref{sec:supertype} provides the details of an algorithm
for calculating the least common supertype of two given types.}

@@

\>[B]{}\mathkw{data}\;\D{\_\sqcup\_\sim\_}\;\mathkw{where}{}\<[E]%

@@

\>[3]{}\Con{lub0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t}\;\D{\sim}\;\V{t}{}\<[E]%

@@

\>[3]{}\Con{lub1}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%

@@

\>[3]{}\Con{lub4}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\Con{nil}\;\D{\sqcup}\;(\Con{list}\;\V{t})\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%

@@

\>[3]{}\Con{lub2}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%

@@

\>[11]{}(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;\V{t'}{}\<[E]%

@@

\>[3]{}\Con{lub2b}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%

@@

\>[12]{}(\Con{listcons}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;\V{t'}{}\<[E]%

@@

\>[3]{}\Con{lub3}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%

@@

\>[13]{}(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{\sim}\;(\Con{list}\;\V{t'}){}\<[E]%

@@

\>[3]{}\Con{lub5}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{t})\;\D{\sqcup}\;\Con{nil}\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%

@@

\>[3]{}\Con{lub6}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\Con{nil}\;\D{\sqcup}\;(\Con{listcons}\;\V{t})\;\D{\sim}\;(\Con{list}\;\V{t}){}\<[E]%

@@

\>[3]{}\Con{lub7}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t1}\;\D{\sim}\;\V{t'}\;\rightarrow\;{}\<[E]%

@@

\>[13]{}(\Con{listcons}\;\V{t})\;\D{\sqcup}\;(\Con{listcons}\;\V{t1})\;\D{\sim}\;(\Con{listcons}\;\V{t'}){}\<[E]%

@@

\>[B]{}\F{\sqcup\textrm{-}comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcup}\;\V{t1}\;\D{\sim}\;\V{t3}{}\<[E]%

@@

\>[B]{}\F{\sqcup\textrm{-}subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%

@@

\>[B]{}\F{\sqcup\textrm{-}least}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{\sim}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%

@@

Since we represent programs using a intrinsically-typed syntax, scope and

@@

natural number with a list as follows. Either the number correspond to some list position,
in which it is of the form \ensuremath{\F{index}\;\V{p}} for some proof \ensuremath{\V{p}\;\mathbin{:}\;\V{x}\;\D{\in}\;\V{xs}} or it is invalid position.

@@

\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\{\mskip1.5mu \V{\Pi}\;\mathbin{:}\;\D{PCtx}\;\V{n}\mskip1.5mu\}\;\{\mskip1.5mu \V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'}\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%

@@

\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}seq}\;\V{p}\;\V{p₁})\;\mathrel{=}\;\Con{seq}\;(\F{forge\textrm{-}types\textrm{-}instr}\;\V{p})\;(\F{forge\textrm{-}types\textrm{-}instr}\;\V{p₁}){}\<[E]%

@@

\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}list}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x}\;\V{x₁})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{idx})\;\V{l}{}\<[E]%

@@

\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}listcons}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x}\;\V{x₁})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{idx})\;\V{l}{}\<[E]%

@@

\subsection{Modifications in the intepreter}\label{sec:changesintep}

>>>>

\begin{abstract}
Formal models are important tools in the programming language research
community. However, such models are full of intricacies and, due to that,
they are subject to subtle errors. Such failures motivated the usage of
tools to ensure the correctness of these formalisms. One way to eliminate
such errors is to encode models in a dependently-typed language in order
to ensure its ``correctness-by-construction''. In this paper, we use this
idea to build a verified interpreter for the list-machine benchmark in the
Agda programming language, comparing the results with formalizations developed
by Appel et. al. We formalize the 14 tasks of the benchmark
using roughly 14\% of LOC compared to a Twelf solution, and 47\% of LOC
compared to a Coq solution, even without the use of proof automation. We also
describe a solution to the second version of the benchmark and compare it with
Appel's et. al. Coq-based solution.
\end{abstract}

@@

In this work, we extend our previous SBLP 2020 paper~\cite{Feitosa2020} by including support to
indirect jumps, as specified in the version 2.0 of the list
machine benchmark~\cite{AppelDL12}. We provide detailed explanations of the needed
changes on the machine syntax, type system and interpreter to accommodate this extension, which
allows the handling of more complex control flow structures, like the call/return model of function calls.

@@

\begin{itemize}
  \item We show how all the details of the list-machine type system
        can be encoded in a dependently-typed syntax setting, which avoids, by construction,
        the presence of stuck states in its definitional interpreter.
  \item We provide a provably correct implementation for testing the subtyping
        relation and to calculate the least common supertype of two input
        types for the machine registers.
  \item We provide a correct-by-construction implementation of a type-checker for the
        list-machine programs. Our type-checker produces, as a result, an intrinsically-typed
        representation of the machine code.
  \item We compare the results of an intrinsic approach encoded in Agda with extrinsic
        formalizations encoded in Twelf and Coq. We show that such intrinsic encoding avoids
        unnecessary repetitions and provides some properties for free.
  \item We provide a detailed discussion on the necessary modifications to the type system,
        type checker and interpreter to support indirect jumps.
\end{itemize}

@@

Dependently typed pattern matching is built by using the so-called
\ensuremath{\mathkw{with}} construct, that allows for matching intermediate
values~\cite{McBride2004}. If the matched value has a dependent type,
then its result can affect the form of other values. For example,
consider the following code that defines a type for natural number
parity. If the natural number is even, it can be represented as the
sum of two equal natural numbers; if it is odd, it is equal to one
plus the sum of two equal values. Pattern matching on a value of
\ensuremath{\D{Parity}\;\V{n}} allows to discover if $n = j + j$ or $n = S (k + k)$,
for some $j$ and $k$ in each branch of \ensuremath{\mathkw{with}}.  Note that the
value of $n$ is specialized accordingly, using information ``learned''
by the type-checker.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{4}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{35}{@{}>{\hspre}l<{\hspost}@{}}%
\column{43}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathkw{data}\;\D{Parity}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\Con{Even}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\;\mathbin{:}\;\D{\mathbb{N}}\mskip1.5mu\}\;\to \;\D{Parity}\;(\V{n}\;\F{+}\;\V{n}){}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\Con{Odd}\;{}\<[9]%
\>[9]{}\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\;\mathbin{:}\;\D{\mathbb{N}}\mskip1.5mu\}\;\to \;\D{Parity}\;(\V{S}\;(\V{n}\;\F{+}\;\V{n})){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\F{parity}\;\mathbin{:}\;(\V{n}\;\mathbin{:}\;\D{\mathbb{N}})\;\to \;\D{Parity}\;\V{n}{}\<[E]%
\\
\>[B]{}\F{parity}\;\mathrel{=}\orange{\texttt{-- definition omitted}}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\F{natToBin}\;\mathbin{:}\;\D{\mathbb{N}}\;\to \;\D{List}\;\D{Bool}{}\<[E]%
\\
\>[B]{}\F{natToBin}\;\Con{zero}\;\mathrel{=}\;\Con{\lbrack\:\rbrack}{}\<[E]%
\\
\>[B]{}\F{natToBin}\;\V{k}\;\mathkw{with}\;(\F{parity}\;\V{k}){}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\F{natToBin}\;(\V{j}\;\F{+}\;\V{j})\;{}\<[25]%
\>[25]{}\mid \;\Con{Even}\;\mathrel{=}\;\Con{false}\;\Con{::}\;\F{natToBin}\;\V{j}{}\<[E]%
\\
\>[B]{}\hsindent{4}{}\<[4]%
\>[4]{}\F{natToBin}\;(\Con{suc}\;(\V{j}\;\F{+}\;\V{j}))\;\mid \;\Con{Odd}\;{}\<[35]%
\>[35]{}\mathrel{=}\;\Con{true}\;{}\<[43]%
\>[43]{}\Con{::}\;\F{natToBin}\;\V{j}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
A important application of Agda's dependently-typed pattern matching is
to rewrite an equality. As an example, consider the following code which
postulates the commutativity of natural number addition and a predicate
over natural numbers, \ensuremath{\D{P}}.

@@

An important application of dependent types is to encode programming languages
syntax. The role of dependent types in this domain is to encode programs that
only allow well-typed and well-scoped terms~\cite{Benton2012}. Intuitively, we encode
the static semantics of the object language in the host language AST's
constructor, leaving the responsibility of checking type safety to the
host language type-checker. Before considering our solution to encode list-machine
programs as intrinsically-typed syntax, we review this approach in a reduced example.

@@

Each program variable is assigned to a list type (Figure~\ref{fig:type-syntax}), which is used to guarantee the safety of executing
fetch-field instructions that demands non-empty list arguments. In order to express such refinements, types are subject to a
subtyping relation. The meta-variable $\tau$ denotes an arbitrary type.
\begin{figure}[H]
\[
\begin{array}{rcll}
  \tau & ::=  & \text{nil} & \text{(type for empty lists)}\\
       & \mid & \text{list }\tau & \text{(lists whose elements have type $\tau$)}\\
       & \mid & \text{listcons }\tau & \text{(non-empty lists of $\tau$)}\\
\end{array}
\]
\centering
\caption{Type syntax}
\label{fig:type-syntax}
\end{figure}
The notation $\tau \subset \tau'$
denotes the subtyping judgment, which is defined as follows.
\begin{figure}[H]
\[
\begin{array}{c}
  \inference{}
            {\tau\subset \tau}
            [subtype-refl]
  \\ \\
  \inference{}
            {nil \subset list\:\tau}
            [subtype-nil]
  \\ \\
  \inference{\tau \subset \tau'}
            {list\:\tau\subset list\:\tau'}
            [subtype-list]\\ \\
  \multicolumn{3}{c}{
  \inference{\tau \subset \tau'}
            {listcons\:\tau\subset list\:\tau'}
            [subtype-listmixed]} \\ \\
  \multicolumn{3}{c}{
            \inference{\tau \subset \tau'}
            {listcons\:\tau\subset listcons\:\tau'}
            [subtype-listcons]} \\ \\
\end{array}
\]
\centering
\caption{Subtyping relation}
\label{fig:subtyping}
\end{figure}
Basically, the subtyping relation specifies that $nil$ (empty list type) is
subtype of any list type and $listcons\:\tau$ is a subtype of the $list\:\tau$.
The other rules specify that type constructors $list$ and $listcons$ respect
the subtyping relation. The least common supertype $\tau = \tau_1 \sqcup \tau_2$ of
$\tau_1$ and $\tau_2$ is defined as the smallest type such that $\tau_1$ and $\tau_2$
are subtypes of $\tau$.

@@

\section{Intrinsically-Typed Syntax}\label{sec:typing}

@@

\>[26]{}\rightarrow\;\V{\tau_1}\;\D{\in}\;\V{\Gamma}\;\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcup}\;\V{\tau_1}\;\D{=}\;\Con{list}\;\V{\tau}\;{}\<[E]%

@@

\>[26]{}\rightarrow\;\Con{list}\;\V{\tau_0}\;\D{\sqcup}\;\V{\tau_1}\;\D{=}\;\Con{list}\;\V{\tau}\;{}\<[E]%

@@

\paragraph{List construction}{The \ensuremath{\Con{instr\textrm{-}cons\textrm{-}new}} and \ensuremath{\Con{instr\textrm{-}cons\textrm{-}upd}} constructors are
used to create a new list. The first creates a new variable, and the second updates an
existing variable. The list is created from two variables, \ensuremath{\V{\tau_0}\;\D{\in}\;\V{\Gamma}} and \ensuremath{\V{\tau_1}\;\D{\in}\;\V{\Gamma}},
which are represented as \emph{de Bruijn} indices. The type of the new list is defined by
the least common supertype\footnote{A complete explanation about the least common supertype
can be found in the original list-machine paper~\cite{Appel07}.}, which is defined by the
type \ensuremath{\V{\tau_1}\;\D{\sqcup}\;\V{\tau_2}\;\D{=}\;\V{\tau}}, which encodes that the least common supertype of \ensuremath{\V{\tau_1}} and \ensuremath{\V{\tau_2}} is \ensuremath{\V{\tau}}.
The resulting typing context adds information about the newly created list.
Subsection~\ref{sec:supertype} provides the details of an algorithm
for calculating the least common supertype of two given types.}

@@

\>[B]{}\mathkw{data}\;\D{\_\sqcup\_=\_}\;\mathkw{where}{}\<[E]%

@@

\>[3]{}\Con{lub0}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t}\;\D{=}\;\V{t}{}\<[E]%

@@

\>[3]{}\Con{lub1}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;\Con{nil}\;\D{=}\;(\Con{list}\;\V{t}){}\<[E]%

@@

\>[3]{}\Con{lub4}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\Con{nil}\;\D{\sqcup}\;(\Con{list}\;\V{t})\;\D{=}\;(\Con{list}\;\V{t}){}\<[E]%

@@

\>[3]{}\Con{lub2}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{=}\;\V{t'}\;\rightarrow\;{}\<[E]%

@@

\>[11]{}(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{listcons}\;\V{t1})\;\D{=}\;\V{t'}{}\<[E]%

@@

\>[3]{}\Con{lub2b}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{=}\;\V{t'}\;\rightarrow\;{}\<[E]%

@@

\>[12]{}(\Con{listcons}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{=}\;\V{t'}{}\<[E]%

@@

\>[3]{}\Con{lub3}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t1}\;\D{=}\;\V{t'}\;\rightarrow\;{}\<[E]%

@@

\>[13]{}(\Con{list}\;\V{t})\;\D{\sqcup}\;(\Con{list}\;\V{t1})\;\D{=}\;(\Con{list}\;\V{t'}){}\<[E]%

@@

\>[3]{}\Con{lub5}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;(\Con{listcons}\;\V{t})\;\D{\sqcup}\;\Con{nil}\;\D{=}\;(\Con{list}\;\V{t}){}\<[E]%

@@

\>[3]{}\Con{lub6}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\mskip1.5mu\}\;\rightarrow\;\Con{nil}\;\D{\sqcup}\;(\Con{listcons}\;\V{t})\;\D{=}\;(\Con{list}\;\V{t}){}\<[E]%

@@

\>[3]{}\Con{lub7}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{t}\;\V{t1}\;\V{t'}\mskip1.5mu\}\;\rightarrow\;\V{t}\;\D{\sqcup}\;\V{t1}\;\D{=}\;\V{t'}\;\rightarrow\;{}\<[E]%

@@

\>[13]{}(\Con{listcons}\;\V{t})\;\D{\sqcup}\;(\Con{listcons}\;\V{t1})\;\D{=}\;(\Con{listcons}\;\V{t'}){}\<[E]%

@@

\>[B]{}\F{\sqcup\textrm{-}comm}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{=}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{\sqcup}\;\V{t1}\;\D{=}\;\V{t3}{}\<[E]%

@@

\>[B]{}\F{\sqcup\textrm{-}subtype}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{=}\;\V{t3}\;\rightarrow\;(\V{t1}\;\D{<:}\;\V{t3})\;\D{\land}\;(\V{t2}\;\D{<:}\;\V{t3}){}\<[E]%

@@

\>[B]{}\F{\sqcup\textrm{-}least}\;\mathbin{:}\;\V{t1}\;\D{\sqcup}\;\V{t2}\;\D{=}\;\V{t3}\;\rightarrow\;\V{t1}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t2}\;\D{<:}\;\V{t3}\;\rightarrow\;\V{t3}\;\D{<:}\;\V{t4}{}\<[E]%

@@

Since we represent programs using an intrinsically-typed syntax, scope and

@@

natural number with a list as follows. Either the number corresponds to some list position,
in which it is of the form \ensuremath{\F{index}\;\V{p}} for some proof \ensuremath{\V{p}\;\mathbin{:}\;\V{x}\;\D{\in}\;\V{xs}} or it is an invalid position.

@@

\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{n}\mskip1.5mu\}\;\{\mskip1.5mu \V{\Pi}\;\mathbin{:}\;\D{PCtx}\;\V{n}\mskip1.5mu\}\;\{\mskip1.5mu \V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'}\;\rightarrow\;\D{UInstr}\;\V{n}{}\<[E]%

@@

\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}seq}\;\V{p}\;\V{p₁})\;\mathrel{=}\;\Con{seq}\;(\F{forget\textrm{-}types\textrm{-}instr}\;\V{p})\;(\F{forget\textrm{-}types\textrm{-}instr}\;\V{p₁}){}\<[E]%

@@

\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}list}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x}\;\V{x₁})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{idx})\;\V{l}{}\<[E]%

@@

\>[B]{}\F{forget\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}branch\textrm{-}listcons}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x}\;\V{x₁})\;\mathrel{=}\;\Con{branch\textrm{-}if\textrm{-}nil}\;(\F{index}\;\V{idx})\;\V{l}{}\<[E]%

@@

\subsection{Modifications in the interpreter}\label{sec:changesintep}
